---
title: "Population_Weighted_ZIP_Centers"
output: html_document
date: "2026-01-21"
---

This calculates population-weighted ZIP code centroids for the entire United States using:

- 2020 Decennial Census** tract-level population data
- HUD Q2 2020 Tract-ZIP Crosswalk (Q2 2020)

## Why Population-Weighted Centroids?

Problem with geometric centroids:
- Geographic center assumes even population distribution
- In reality, people cluster in neighborhoods
- Suburban ZIP might have geometric center in empty field

Solution:
- Weight each Census tract's location by its population
- Pull centroid toward where people actually live
- More accurate for accessibility analysis than blocks or block groups as they can
be distributed over the state or county boundries. 

```{r load-packages}

setwd("~/Desktop/THESIS_ANALYSIS/Heatmaps/ZIP_CENTERS_TRACT_TO_ZIP")
# Spatial data and mapping
library(tidycensus)   # Census data API
library(tigris)       # Census geographic boundaries
library(sf)           # Spatial operations
library(dplyr)        
library(purrr)        
library(readxl)       
library(janitor)
library(ggplot2)      
library(scales)      
options(tigris_use_cache = TRUE)  # Cache geographic downloads
```

**Crosswalk variables**

zip: USPS ZIP code
tract: Census tract GEOID (geographic identifiers)
res_ratio: share of residential addresses in the tract that fall in this ZIP

ZIP codes are padded to 5 digits to prevent merge failures
Only residential overlaps (`res_ratio > 0`) are retained


```{r}
crosswalk_file <- "TRACT_ZIP_062020.xlsx"

tract_zip_crosswalk <- read_excel(crosswalk_file) %>%
  clean_names() %>%
  select(zip, tract, res_ratio) %>%
  mutate(
    tract = as.character(tract),
    zip   = sprintf("%05d", as.numeric(zip))
  ) %>%
  filter(res_ratio > 0)
```


Key variables:
- `res_ratio`: Proportion of tract's residential addresses in this ZIP
- `bus_ratio`: Proportion of tract's business addresses in this ZIP
- We use `res_ratio` because we're weighting by residential population
## Download Census Tract Populations

**Source:** 2020 Decennial Census  
**Variable:** P1_001N (Total Population)  
**Geography:** Census Tracts  

```{r census}

rds <- "us_tracts_2020.rds"

if (file.exists(rds)) {
  tracts <- readRDS(rds)
} else {
  tracts <- map_dfr(c(state.abb, "DC"), function(s) {
    cat(s, "")
    get_decennial(
      geography = "tract",
      variables = "P1_001N",
      state = s,
      year = 2020,
      geometry = TRUE
    )
  })
  saveRDS(tracts, rds)
}

```

~ Only tracts appearing in the tract–ZIP crosswalk are retained
~ Zero-population tracts are excluded
~ Tracts are projected to EPSG:5070 to enable population-weighted spatial
  averaging in meters

```{r filter}

tracts_proj <- tracts %>%
  filter(GEOID %in% tract_zip_crosswalk$tract, value > 0) %>% 
  st_transform(5070)

```


**Tract Representative Points (Point-on-Surface)**

- Converts each census tract polygon into a single representative point inside
  the tract
- Uses point-on-surface to ensure the point lies within tract boundaries
- Extracts x and y coordinates (in meters) from each representative point
- Attaches tract population to each point


```{r}
tract_points <- tracts_proj %>%
  mutate(
    point = st_point_on_surface(geometry),
    coords = st_coordinates(point),
    x = coords[,1],
    y = coords[,2]
  ) %>%
  st_drop_geometry() %>%
  select(
    tract = GEOID,
    x,
    y,
    pop = value
  )

```

**Population-Weighted ZIP Centroids**

- Merges the tract–ZIP crosswalk with tract representative points and populations
- Drops observations with missing coordinates or zero population
- Groups tracts by ZIP code
- Computes population weights for each tract within a ZIP (pop × res_ratio) and              renormalizes them to sum to one
- Calculates ZIP-level centroid coordinates as weighted averages of tract coordinates
- Computes ZIP-level population and the number of contributing tracts
- Converts results to a spatial point object (EPSG:5070) and then re-projects to             latitude/longitude (EPSG:4326)
- Extracts longitude and latitude 

```{r Population-Weighted ZIP Centroids}

zip_cents <- tract_zip_crosswalk %>%
  left_join(tract_points, by = "tract") %>%       # attach tract coordinates and population
  filter(!is.na(x), pop > 0) %>%                   # keep valid, populated tracts only
  group_by(zip) %>%                                # operate within each ZIP code
  mutate(
    pop_weighted_value = pop * res_ratio,           # population assigned to ZIP
    weight = pop_weighted_value / sum(pop_weighted_value)  # normalize weights (sum to 1)
  ) %>%
  summarize(
    x = sum(x * weight),                            # population-weighted X coordinate
    y = sum(y * weight),                            # population-weighted Y coordinate
    population = sum(pop * res_ratio),              # total ZIP population
    n_tracts = n(),                                 # number of contributing tracts
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("x", "y"), crs = 5070) %>%    # convert to spatial points (meters)
  st_transform(4326) %>%                            # re-project to lat/long
  mutate(
    longitude = st_coordinates(.)[,1],              # extract longitude
    latitude  = st_coordinates(.)[,2]               # extract latitude
  )                


```


We download the ZCTA file
```{r ZCTA}

zcta_boundaries <- zctas(year = 2020, cb = FALSE) %>%
  st_transform(5070)  # Project to meters for accurate area calculation

```


```{r ZCTA Area}
zip_geometry <- zcta_boundaries %>%
  mutate(
    area_sq_m = st_area(geometry),                    # Area in square meters
    area_sq_km = as.numeric(area_sq_m) / 1e6,        # Convert to square kilometers
    area_sq_mi = as.numeric(area_sq_m) / 2.589988e6  # Convert to square miles
  ) %>%
  select(
    zip = ZCTA5CE20,  # ZIP code identifier (adjust column name if different)
    area_sq_km,
    area_sq_mi
  ) %>%
  st_drop_geometry()  # Remove geometry, keep just metrics
```


The st_bbox() functionis used to return the bounding box of a simple feature or a set of simple features. It computes the minimum and maximum coordinates (xmin, ymin, xmax, ymax) that enclose all the geometries within the object. 

```{r ZIP length}
zip_length <- zcta_boundaries %>%
  st_transform(5070) %>%  # Ensure projected CRS for accurate distances
  mutate(
    length_km = map_dbl(geometry, function(g) {
      bbox <- st_bbox(g)
      width <- bbox["xmax"] - bbox["xmin"]
      height <- bbox["ymax"] - bbox["ymin"]
      diagonal <- sqrt(width^2 + height^2)
      diagonal / 1000  # Convert meters to km
    }),
    length_mi = length_km * 0.621371
  ) %>%
  select(
    zip = ZCTA5CE20,
    length_km,
    length_mi
  ) %>%
  st_drop_geometry()
```


```{r merge}

zip_cents_complete <- zip_cents %>%
  left_join(zip_geometry, by = "zip") %>%
  left_join(zip_length, by = "zip") %>%
  select(
    zip,
    longitude,
    latitude,
    population,
    n_tracts,
    area_sq_km,
    area_sq_mi,
    length_km,
    length_mi,
  ) 
```


```{r csv}
write.csv(zip_cents_complete,"Pop_Centroid.csv")

```


```{r}
zip_cents_complete1 <- zip_cents %>%
  left_join(zip_geometry, by = "zip") %>%
  left_join(zip_length, by = "zip") %>%
  select(
    zip,
    longitude,
    latitude,
    population,
    n_tracts,
    area_sq_km,
    area_sq_mi,
    length_km,
    length_mi,
  ) 
```


```{r shapefile}
st_write(
  zip_cents_complete1,
  "zip_centroid.shp",
  delete_dsn = TRUE
)

```


